Testing and Isolation Efficacy: Insights from a Simple Epidemic Model

### 2021-06-04 SIR_manuscript.tex

\begin{table}[htp]
\centering
\begin{tabular}{|c|L{2in}cc|} \hline
  Symbol & Description & Unit & Value \\ \hline
  $N_0$     & Total population size & people & $10^6$ \\ \hline
  $\omega$  & Rate of test return, i.e., rate of onward flow from ``waiting'' to ``confirmed'' or ``untested'' compartments  & 1/day & - \\ \hline
  $\gamma$ & Recovery rate & 1/day & 1/6 \\ \hline 
  $\rho$   & \percap testing intensity & 1/day & - \\ \hline 
  $\theta\_w$ & Isolation efficacy (reduction of the transmission probability) for ``waiting'' individuals & - & - \\ \hline
  $\theta\_c$ & Isolation efficacy for ``confirmed positive'' individuals & - & -  \\ \hline
  $\beta$ & Transmission rate & 1/day & 0.25 \\ \hline
  $\Lambda$ & Force of infection & 1/day & - \\ \hline
  $p_S$ & Probability of positive tests for $S$ ($= 1-\textrm{specificity}$) & - & 0 \\ \hline
  $p_I$ & Probability of positive tests for $I$ ($= \textrm{sensitivity}$) & - & 1 \\ \hline
  $p_R$ & Probability of positive tests for $R$ ($= 1-\textrm{specificity}$) & - & 0.5 \\ \hline
  $w_S, w_I, w_R$ & Relative testing weights & - &
  \begin{minipage}[t]{0.21\columnwidth}%
    Random: $\{1,1,1\}$ \\ Targeted: $\{0.3,1,1\}$
  \end{minipage} \\
  \hline
  \end{tabular}
\caption{\label{tab:params} Parameters of the model \eqref{model}.}
\end{table}

@@

The low-testing case \fref{pan} reflects the case where $rho$ is small relative to the population size. Specifically, $\rho \in [0,0.013]$, and test return rate $\omega\in [1/12,2]$. This testing intensity reflects realistic testing rates during the \covid pandemic, i.e. a maximum of 1.3\% of the population per day, giving a maximum of 10000 tests/day in a population of size $N_0=10^6$ (this is approximately four times the maximum testing rate in Ontario, Canada in mid-2021). The less realistic high-testing case \fref{pan2} is included to highlight the occurrence of non-monotonic changes in $\Rnum$ with respect to $\rho$.
In \fref{pan2} the maximum capacity of $\rho$ is larger relative to the population size, $\rho \in [0,1/5)$ and the test return rate $\omega\in [1/5,2]$; these values are clearly unrealistic for a large population but might be relevant for some population undergoing focused testing, such as a sports league or university.
  The critical contour of $\Delta=1-\frac{\gamma}{\beta}$, corresponding to the threshold $\Rnum=1$, is shown as a dotted line in \fref{pan} and \fref{pan2}; combinations of $\rho$ and $\omega$ that lie above this critical contour will control the epidemic. The implied baseline value of $\Rnum=\frac{\beta}{\gamma}$ in these figures is $\approx 1.5$, corresponding to a case where the epidemic spreads slowly even in the absence of testing and tracing.

@@

\newpage
\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_random.pdf}
\caption{}\label{p.a}
\end{subfigure}
%
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_TTI.pdf}
\caption{}\label{p.b}
\end{subfigure}
\caption{
{\bf A comparison of effectiveness control on $\Rnum$ at low level of \percap testing intensity $\rho$.}
Numerical evaluation of the effectiveness control parameter, $\Delta$ (Eq.~\ref{eq:del4}), with respect to changes in testing and isolation. Parameter values (Table~\ref{tab:params}):
$\beta=0.25$ day$^{-1}$, $1/\gamma= 1/3$ days; $\omega \in [1/12,2]$ day$^{-1}$;  $\rho \in [0,0.013]$ day$^{-1}$; $\theta\_w$ and $\theta\_c$ vary between 0 (no effect of isolation) and 1 (complete elimination of transmission); $p_S=0$, $p_I=1$ and $p_R=0.5$. Contours of $\Delta$ are plotted for (a) random testing ($w_S=w_I=w_R=1$) and (b) targeted testing ($w_S=0.3$; $w_I=w_R=1$). 
The baseline basic reproduction number $\Rnum$ and the doubling time $r$, i.e. $\beta/\gamma$ and $\beta-\gamma$ respectively, are approximately 1.5 and 0.08 in this simulation.
}
\label{pan}
\end{figure}

\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_random2.pdf}
\caption{}
\end{subfigure}
%
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_TTI2.pdf}
\caption{}
\end{subfigure}
\caption{
{\bf A comparison of effectiveness control on $\Rnum$ at high level of \percap testing intensity $\rho$.}
We numerically evaluate the effectiveness control parameter, $\Delta$ (Eq.~\ref{eq:del4}), with respect to testing and isolation. We use the following parameters (see Table~\ref{tab:params} for a brief description):
$\omega \in [1/5,2]$ 1/day, $1/\gamma= 6$ days, $\rho \in [0,1/5)$ 1/day, $\theta\_w$ and $\theta\_c$ vary between 0 and 1 with 0 for no effect and 1 for full effect of isolation on the transmission probability, $\beta=0.25$ 1/day, $p_S=0$, $p_I=1$ and $p_R=0.5$. Contours of $\Delta$ are plotted for two testing strategies identified by a set of relative testing weights; (a) random testing where $w_S=w_I=w_R=1$ and (b) targeted testing where $w_S=0.3$ and $w_I=w_R=1$. The baseline basic reproduction number $\Rnum$ and the doubling time $r$, i.e. $\beta/\gamma$ and $\beta-\gamma$ respectively, are approximately 1.5 and 0.08 in this simulation.
}
\label{pan2}
\end{figure}

@@

\item Lastly, we show that increasing the focus of testing from random to targeted, always increases $\Delta$, thus decreases \Rnum. This can be seen in \fref{pan} and \fref{pan2}. In particular, we realize higher $\Delta$ levels, i.e., darker colors in corresponding plots\/facets as we go from panel~(a) to panel~(b) in the figures. \todo{Ali to restructure or delete; feels redundant.}
\end{itemize}

>>>>

\begin{table}[htp]
\centering
\begin{tabular}{|c|L{2in}cc|} \hline
  Symbol & Description & Unit & Value \\ \hline
  $N_0$     & Total population size & people & $10^6$ \\ \hline
  $\omega$  & Rate of test return, i.e., rate of onward flow from ``waiting'' to ``confirmed'' or ``untested'' compartments  & 1/day & - \\ \hline
  $\gamma$ & Recovery rate & 1/day & 1/6 \\ \hline 
  $\rho$   & \percap testing intensity & 1/day & - \\ \hline 
  $\theta\_w$ & Isolation efficacy (reduction of the transmission probability) for ``waiting'' individuals & - & - \\ \hline
  $\theta\_c$ & Isolation efficacy for ``confirmed positive'' individuals & - & -  \\ \hline
  $\beta$ & Transmission rate & 1/day & 0.5 \\ \hline
  $\Lambda$ & Force of infection & 1/day & - \\ \hline
  $p_S$ & Probability of positive tests for $S$ ($= 1-\textrm{specificity}$) & - & 0 \\ \hline
  $p_I$ & Probability of positive tests for $I$ ($= \textrm{sensitivity}$) & - & 1 \\ \hline
  $p_R$ & Probability of positive tests for $R$ ($= 1-\textrm{specificity}$) & - & 0.5 \\ \hline
  $w_S, w_I, w_R$ & Relative testing weights & - &
  \begin{minipage}[t]{0.21\columnwidth}%
    Random: $\{1,1,1\}$ \\ Targeted: $\{0.3,1,1\}$
  \end{minipage} \\
  \hline
  \end{tabular}
\caption{\label{tab:params} Parameters of the model \eqref{model}.}
\end{table}

@@

The low-testing case \fref{pan} reflects the case where $rho$ is small relative to the population size. Specifically, $\rho \in [0,0.013]$, and test return rate $\omega\in [1/12,2]$. This testing intensity reflects realistic testing rates during the \covid pandemic, i.e. a maximum of 1.3\% of the population per day, giving a maximum of 10000 tests/day in a population of size $N_0=10^6$ (this is approximately four times the maximum testing rate in Ontario, Canada in mid-2021). The less realistic high-testing case \fref{pan2} is included to highlight the occurrence of non-monotonic changes in $\Rnum$ with respect to $\rho$.
In \fref{pan2} the maximum capacity of $\rho$ is larger relative to the population size, $\rho \in [0,1/5)$ and the test return rate $\omega\in [1/5,2]$; these values are clearly unrealistic for a large population but might be relevant for some population undergoing focused testing, such as a sports league or university.
  The critical contour of $\Delta=1-\frac{\gamma}{\beta}$, corresponding to the threshold $\Rnum=1$, is shown as a dotted line in \fref{pan} and \fref{pan2}; combinations of $\rho$ and $\omega$ that lie above this critical contour will control the epidemic. The implied baseline value of $\Rnum=\frac{\beta}{\gamma}$ in these figures is 3, corresponding to a case where the epidemic spreads slowly even in the absence of testing and tracing.

@@

\newpage
\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_random.pdf}
\caption{}\label{p.a}
\end{subfigure}
%
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_TTI.pdf}
\caption{}\label{p.b}
\end{subfigure}
\caption{
{\bf A comparison of effectiveness control on $\Rnum$ at low level of \percap testing intensity $\rho$.}
Numerical evaluation of the effectiveness control parameter, $\Delta$ (Eq.~\ref{eq:del4}), with respect to changes in testing and isolation. Parameter values (Table~\ref{tab:params}):
$\beta=0.5$ 1/day, $1/\gamma= 6$ days; $\omega \in [1/12,2]$ 1/day;  $\rho \in [0,0.013]$ 1/day per capita; $\theta\_w$ and $\theta\_c$ vary between 0 (no effect of isolation) and 1 (complete elimination of transmission); $p_S=0$, $p_I=1$ and $p_R=0.5$. Contours of $\Delta$ are plotted for (a) random testing ($w_S=w_I=w_R=1$) and (b) targeted testing ($w_S=0.3$; $w_I=w_R=1$). 
The baseline basic reproduction number $\Rnum$ and the doubling time $r$, i.e. $\beta/\gamma$ and $\beta-\gamma$ respectively, are 3 and about 0.3 in this simulation.
}
\label{pan}
\end{figure}

\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_random2.pdf}
\caption{}
\end{subfigure}
%
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_TTI2.pdf}
\caption{}
\end{subfigure}
\caption{
{\bf A comparison of effectiveness control on $\Rnum$ at high level of \percap testing intensity $\rho$.}
We numerically evaluate the effectiveness control parameter, $\Delta$ (Eq.~\ref{eq:del4}), with respect to testing and isolation. We use the following parameters (see Table~\ref{tab:params} for a brief description):
$\omega \in [1/5,2]$ 1/day, $1/\gamma= 6$ days, $\rho \in [0,1/5)$ 1/day, $\theta\_w$ and $\theta\_c$ vary between 0 and 1 with 0 for no effect and 1 for full effect of isolation on the transmission probability, $\beta=0.5$ 1/day, $p_S=0$, $p_I=1$ and $p_R=0.5$. Contours of $\Delta$ are plotted for two testing strategies identified by a set of relative testing weights; (a) random testing where $w_S=w_I=w_R=1$ and (b) targeted testing where $w_S=0.3$ and $w_I=w_R=1$. The baseline basic reproduction number $\Rnum$ and the doubling time $r$, i.e. $\beta/\gamma$ and $\beta-\gamma$ respectively, are 3 and 0.3 in this simulation.
}
\label{pan2}
\end{figure}

@@

\end{itemize}
